Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex

\section{Related Work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}

\subsection{Program Calculation}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

@@

\subsection{Termination Checking}

@@

\subsubsection{Sized Types}

>>>>

\section{Related Work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
Finally, we compare our encoding of recursion schemes to other forms of
termination checking (Section~\ref{sec:termcheck}, and sized types
(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
recursion.

\subsection{Program Calculation}\label{sec:calccoq}

Within the domain of program optimization, program calculation serves as a
well-established programming technique aimed at deriving efficient programs
from their naive counterparts through systematic program
transformation~\cite{gibbonsSquiggol}. This area has been extensively explored
over the years. Tesson et al.\ demonstrated the efficacy of leveraging Coq to
establish an approach for implementing a robust system dedicated to verifying
the correctness of program transformations for functions that manipulate
lists~\cite{10.1007/978-3-642-17796-5_10}. Murata and Emoto went further and
formalised recursion schemes in Coq~\cite{MurataE19}. Their development does
not include hylomorphisms and dynamorphisms, and relies on the functional
extensionality axiom, as well as further extensionality axioms for each
coinductive datatype that they use. They do not discuss the extracted OCaml
code from their formalisation.  Larchey-Wendling and Monin encode recursion
schemes in Coq, by formalising computational graphs of
algorithms~\cite{larchey2022braga}. Their work does not focus on encoding
generic recursion schemes, and proving their algebraic laws.  Castro-Perez et
al.~\cite{farmsCastro} encode the laws of hylomorphisms as part of the type
system of a functional language to calculate parallel programs from
specifications. Their work focuses on parallelism, and they do not formalise
their approach in a proof assistant, and the laws of hylomorphisms are axioms
in their system.

@@

\subsection{Termination Checking}\label{sec:termcheck}

@@

\subsubsection{Divide and Conquer Recursion}\label{sec:reccoq}

\subsubsection{Sized Types}\label{sec:sizedT}
